Hierarchical Bayesian regression modelling predicts biochemical rection energies and facilitates quantitative modelling of cell metabolism

### 2023-10-27 paper/main.qmd
Update paper

In the context of kinetic modelling, Bayesian inference is appealing because
it allows uncertainty to be represented appropriately without sacrificing
mechanistic accuracy. Measurement uncertainty can naturally be represented in a
Bayesian measurement model, whereas the prior model can represent quantitative
uncertainty about kinetic parameters. Finally, kinetic rate laws can be
represented in Bayesian data generation models with arbitrarily high fidelity.
See (REFERENCES) for more about Bayesian inference and (REFERENCE) for a
discussion of practical Bayesian workflow.

@@

The simulated dataset and underlying kinetic model that we used for our analysis
can be found in [#Supplementary Info].

@@

@@

We used these parameters to simulate steady states in a range of plausible
experimental conditions, again using @saa_general_2015 as a starting point.
These steady states were then used to generate simulated measurements using the
measurement model described below.

\begin{equation}
a = b \label{eq-meth-ode}
\end{equation}

@@

We constructed inputs in Maud’s format for each of the analysed datasets, based on the scenario that the true kinetic model was known except for parameter values, which needed to be inferred from the training data and priors. These inputs can be found at XXXX.

The prior distributions and corresponding true parameter values used in our case
study are shown in [Supplementary]. The first two columns show the 1% and 99%
quantiles of each marginal prior distribution. True parameter value are shown
in column three, and the last column shows the z-score on log scale of the true
parameter value according the marginal prior distribution. As seen in table
2 [REFERENCE] there are 7 parameters for which the true value is outside the
1%-99% range. This is desirable, making the case study more realistic, because
extreme deviance from the prior distribution is likely to occur in practice due
to in vivo to in vitro measurement differences.

@@

Overall, our results show that Maud can fit a realistic pathway-sized dataset.
This was achieved without fixing the marginal values of kinetic parameters: the
information required to make good predictions was contained in the correlation
structure of the joint posterior distribution. This finding is consistent with
previous analyses of biological systems that found they are “sloppy”, that is,
sensitive to parameter combinations rather than marginal parameter values, with
important combinations, scales and regions of sensitivity being difficult to
ascertain in advance @gutenkunst_2007, @poirier_revising_1998.

The question naturally arises whether the crucial high-dimensional parameter
correlations are linear or non-linear. This is relevant to the question of
model performance, as linear correlations are easier to correct for. A linearly
correlated posterior space would also be easier to summarise. We address this
question below in section XXX.

This case study illustrates the type of kinetic model and dataset that Maud
can fit. The model we analysed has XXX reactions, XXX state variables and XXX
parameters. Posterior sampling with adaptive Hamiltonian Monte Carlo generated
XXX ESS/hour. Generalising from this result, we conclude that it is feasible
to use this method to fit models in the same order of magnitude, but not, for
example, genome-scale kinetic models. To fit larger models, faster steady state
solving methods or alternative inference algorithms will be required. Section
XXX addresses whether Laplace approximation is a suitable candidate.

### Comparison with Laplace approximation

@@

The Laplace approximation yielded XXX samples in XXX time. The diagnostics
indicated that our algorithm was able to find the maximum a posteriori parameter
configuration, approximate the Hessian and use these quantities to generate
approximate posterior samples. The results can be found at XXX.

@@

Maud is a command line application implementing Bayesian inference for
a wide range of realistic kinetic models. Maud is written in Python
[@vanrossumPythonReferenceManual2009], designed for use on Windows, macOS
and Linux, registered on the Python Package Index as `maud-metabolic-models`,
documented at <https://maud- metabolic-models.readthedocs.io> and actively
developed and maintained at <https://github.com/biosustain/Maud/>.

@@

Maud inputs are structured directories, somewhat inspired by the PEtab
format @SchmiesterSch2021. A Maud input directory must contain a toml
[@preston-wernertomandgedampradyunTOMLSpecification0rc2020] file called
`config.toml` which gives the input a name, configures how Maud will be run
and tells Maud where to find the other files, allowing these to have custom
names. It must also include a file containing a kinetic model definition,
a file specifying information about parameters and a file with information
experiments. The required structure of these files is documented at <https://
maud-metabolic-models.readthedocs.io/en/latest/inputting.html>. The input is
validated against a Pydantic [@pydanticdevelopersPydantic2022] data model which can be
inspected at <https://github.com/biosustain/Maud/tree/main/maud/data_model>.

@@

Each of the terms on the right-hand side of (REFER TO EQUATION) is a function
of $C$ and $\theta$. This idea is taken from @noor_note_2013. The terms
usefully gather physically meaningful and conceptually distinct factors
contributing to reaction fluxes. $Enzyme$ captures the effect of enzyme
concentration, $k_{cat}$ that of enzyme efficiency, $Reversibility$ quantifies
thermodynamic effects, $Saturation$ the effect of enzyme availability and
$Allostery$ the effect of post translational modifications.

@@

Maud’s prior model includes unknown parameters corresponding to quantities
in the kinetic model that are assumed to be unknown, other than steady state
metabolite concentrations and fluxes, which are derived from the values of
other parameters by solving the steady state problem. See Table 1 (REFERENCE)
for a description of all these parameters and their dimensions. Note that some
quantities in Maud’s kinetic model are not treated as parameters: for example,
temperatures, compartment volumes and the formation energy of water. Maud
treats these quantities as if they were known precisely: they can be configured
by the user or default values can be used. Although in practice there can be
considerable uncertainty regarding these quantities, we chose to disregard this
uncertainty in the interest of simplicity.

@@

\begin{equation}
y_i^{conc} \sim LN(\ln{\hat{y}_i^{conc}}, \sigma_i^{conc})\label{eq-yconc}
\end{equation}

\begin{equation}
y_i^{enz} \sim LN(\ln{\hat{y}_i^{enz}}, \sigma_i^{enz})\label{eq-yenz}
\end{equation}

Both equations are log-normal generalised linear models with a standard
link function (the natural logarithm $\ln$) and known standard deviation
$ \sigma_{𝑐𝑜𝑛𝑐}$. The use of this measurement model is motivated by the
consideration that concentrations are constrained to be non-negative, so the
measurement model should avoid assigning positive probability mass to negative
metabolite concentration values. In addition, we expect the precision of most
metabolomics and proteomics experiments to be roughly proportional to the value
of the true measured quantity, which supports a measurement model with constant
coefficient of variation. The measurement standard deviations $\sigma_{𝑐𝑜𝑛𝑐}$
and $\sigma_{𝑒𝑛𝑧}$ are assumed to be known exactly for simplicity; plausible
values can be elicited by considering the likely coefficient of variation of the
measuring apparatus.

@@

Finally, the reader may wonder why Maud uses a linear regression model for
reaction flux measurements even though this creates the potential for erroneous
double counting and requires non-trivial upstream modelling, as intracellular
fluxes typically cannot be measured directly, but must be inferred from isotope
labelling experiments using metabolic flux analysis (see REFERENCE). Ideally
Maud’s measurement model for fluxes would extend from fluxes to the results of
potential labelling experiments, thereby removing the need for upstream analysis
and avoiding any double counting. This option has not yet been pursued, again
for the sake of simplicity.

@@

Although integrals of the joint probability model for kinetic models
are typically analytically intractable, they can be approximated
numerically using Markov Chain Monte Carlo (MCMC) and other methods.
Maud uses MCMC primarily because there exist many methods for verifying
that MCMC samples really do approximate the target probability
distribution: see@vehtariRankNormalizationFoldingLocalization2021 and
@taltsValidatingBayesianInference2018 for discussion of this point. In
addition, there are several examples of successful Bayesian kinetic modelling
projects using MCMC including @st.johnBayesianInferenceMetabolic2018 and
@xingModelingKineticsLargescale2010.

>>>>

In the context of kinetic modelling, Bayesian inference is appealing because
it allows uncertainty to be represented appropriately without sacrificing
mechanistic accuracy. Measurement uncertainty can naturally be represented in a
Bayesian measurement model, whereas the prior model can represent quantitative
uncertainty about kinetic parameters. Finally, kinetic rate laws can be
represented in Bayesian data generation models with arbitrarily high fidelity.
See @gelmanBayesianDataAnalysis2020a for more about Bayesian inference and
@gelmanBayesianWorkflow2020 for a discussion of practical Bayesian workflow.

@@

The simulated dataset and underlying kinetic model that we used for our analysis
can be found at <https://github.com/biosustain/Methionine_model/tree/main/data/methionine>
and is described in supplementary information section [REFERENCE].

@@

\begin{align}
\frac{d[met]}{dt} &= v_{Influx} - v_{PROT} - v_{MAT} +v_{MS} + v_{BHMT} \label{eq-meth-ode} \\
\frac{d[amet]}{dt} &= v_{MAT} - v_{GNMT} - v_{METH} \nonumber \\
\frac{d[ahyc]}{dt} &= v_{GNMT} + v_{METH} - v_{AHC} \nonumber \\
\frac{d[hyc-L]}{dt} &= v_{AHC} - v_{CBS} - v_{MS} - v_{BHMT} \nonumber \\
\frac{d[5mthf]}{dt} &= v_{MTHFR} - v_{MS} \nonumber 
\end{align}

@@

We used these parameters to simulate steady states in a range of plausible
experimental conditions, again using @saa_general_2015 as a starting point.
These steady states were then used to generate simulated measurements using the
measurement model described below in @sec-methionine-measurement-model.

### Measurement model {#sec-methionine-measurement-model}

@@

We constructed inputs in Maud’s format for each of the analysed datasets, based
on the scenario that the true kinetic model was known except for parameter
values, which needed to be inferred from the training data and priors. These
inputs can be found at <https://github.com/biosustain/Methionine_model/tree/main/data>.

The prior distributions and corresponding true parameter values used in our
case study are shown in supplementary materials section [REFERENCE]. The first
two columns show the 1% and 99% quantiles of each marginal prior distribution.
True parameter value are shown in column three, and the last column shows the
z-score on log scale of the true parameter value according the marginal prior
distribution. As can be seen from the table, there are 7 parameters for which
the true value is outside the 1%-99% range. This is desirable, making the case
study more realistic, because extreme deviance from the prior distribution is
likely to occur in practice due to in vivo to in vitro measurement differences.

@@

Overall, our results show that Maud can fit a realistic pathway-sized dataset.
This was achieved without fixing the marginal values of kinetic parameters: the
information required to make good predictions was contained in the correlation
structure of the joint posterior distribution. This finding is consistent with
previous analyses of biological systems that found they are “sloppy”, that is,
sensitive to parameter combinations rather than marginal parameter values, with
important combinations, scales and regions of sensitivity being difficult to
ascertain in advance [@gutenkunst_2007; @poirier_revising_1998].

The question naturally arises whether the crucial high-dimensional parameter
correlations are linear or non-linear. This is relevant to the question of
model performance, as linear correlations are easier to correct for. A linearly
correlated posterior space would also be easier to summarise. We address this
question below in @sec-laplace.

This case study illustrates the type of kinetic model and dataset that Maud
can fit. The model we analysed has XXX reactions, XXX state variables and XXX
parameters. Posterior sampling with adaptive Hamiltonian Monte Carlo generated
XXX ESS/hour. Generalising from this result, we conclude that it is feasible
to use this method to fit models in the same order of magnitude, but not,
for example, genome-scale kinetic models. To fit larger models, faster steady
state solving methods or alternative inference algorithms will be required.
@sec-laplace addresses whether Laplace approximation is a suitable candidate.

### Comparison with Laplace approximation {#sec-laplace}

@@

The Laplace approximation yielded XXX samples in XXX time. The diagnostics
indicated that our algorithm was able to find the maximum a posteriori parameter
configuration, approximate the Hessian and use these quantities to generate
approximate posterior samples. The results can be found at <https://github.com/biosustain/Methionine_model/tree/main/results>.

@@

Maud is a command line application implementing Bayesian inference for
a wide range of realistic kinetic models. Maud is written in Python
[@vanrossumPythonReferenceManual2009], designed for use on Windows, macOS
and Linux, registered on the Python Package Index as `maud-metabolic-models`,
documented at <https://maud-metabolic-models.readthedocs.io> and actively
developed and maintained at <https://github.com/biosustain/Maud/>.

@@

Maud inputs are structured directories, somewhat inspired by the PEtab
format [@SchmiesterSch2021]. A Maud input directory must contain a toml
[@preston-wernertomandgedampradyunTOMLSpecification0rc2020] file called
`config.toml` which gives the input a name, configures how Maud will be run
and tells Maud where to find the other files, allowing these to have custom
names. It must also include a file containing a kinetic model definition,
a file specifying information about parameters and a file with information
experiments. The required structure of these files is documented at <https://
maud-metabolic-models.readthedocs.io/en/latest/inputting.html>. The input is
validated against a Pydantic [@pydanticdevelopersPydantic2022] data model which can be
inspected at <https://github.com/biosustain/Maud/tree/main/maud/data_model>.

@@

Each of the terms on the right-hand side of \eqref{eq-decomposition} is a
function of $C$ and $\theta$. This idea is taken from @noor_note_2013. The
terms usefully gather physically meaningful and conceptually distinct factors
contributing to reaction fluxes. $Enzyme$ captures the effect of enzyme
concentration, $k_{cat}$ that of enzyme efficiency, $Reversibility$ quantifies
thermodynamic effects, $Saturation$ the effect of enzyme availability and
$Allostery$ the effect of post translational modifications.

@@

Maud’s prior model includes unknown parameters corresponding to quantities
in the kinetic model that are assumed to be unknown, other than steady state
metabolite concentrations and fluxes, which are derived from the values of
other parameters by solving the steady state problem. See table (REFERENCE) in
this paper's supplementary materials for a description of all these parameters
and their dimensions. Note that some quantities in Maud’s kinetic model are
not treated as parameters: for example, temperatures, compartment volumes and
the formation energy of water. Maud treats these quantities as if they were
known precisely: they can be configured by the user or default values can be
used. Although in practice there can be considerable uncertainty regarding
these quantities, we chose to disregard this uncertainty in the interest of
simplicity.

@@

\begin{align}
y_i^{conc} &\sim LN(\ln{\hat{y}_i^{conc}}, \sigma_i^{conc})\label{eq-yconc} \\
y_i^{enz} &\sim LN(\ln{\hat{y}_i^{enz}}, \sigma_i^{enz})\label{eq-yenz}
\end{align}

Both equations are log-normal generalised linear models with a standard
link function (the natural logarithm $\ln$) and known standard deviation
$\sigma_{𝑐𝑜𝑛𝑐}$. The use of this measurement model is motivated by the
consideration that concentrations are constrained to be non-negative, so the
measurement model should avoid assigning positive probability mass to negative
metabolite concentration values. In addition, we expect the precision of most
metabolomics and proteomics experiments to be roughly proportional to the value
of the true measured quantity, which supports a measurement model with constant
coefficient of variation. The measurement standard deviations $\sigma_{𝑐𝑜𝑛𝑐}$
and $\sigma_{𝑒𝑛𝑧}$ are assumed to be known exactly for simplicity; plausible
values can be elicited by considering the likely coefficient of variation of the
measuring apparatus.

@@

Finally, the reader may wonder why Maud uses a linear regression model for
reaction flux measurements even though this creates the potential for erroneous
double counting and requires non-trivial upstream modelling, as intracellular
fluxes typically cannot be measured directly. Instead, fluxes are typically
inferred from isotope labelling experiments using metabolic flux analysis:
see @daiUnderstandingMetabolismFlux2017 for more about this method. Ideally
Maud’s measurement model for fluxes would extend from fluxes to the results of
potential labelling experiments, thereby removing the need for upstream analysis
and avoiding any double counting. This option has not yet been pursued, again
for the sake of simplicity.

@@

Although integrals of the joint probability model for kinetic models
are typically analytically intractable, they can be approximated
numerically using Markov Chain Monte Carlo (MCMC) and other methods.
Maud uses MCMC primarily because there exist many methods for verifying
that MCMC samples really do approximate the target probability
distribution: see @vehtariRankNormalizationFoldingLocalization2021 and
@taltsValidatingBayesianInference2018 for discussion of this point. In
addition, there are several examples of successful Bayesian kinetic modelling
projects using MCMC including @st.johnBayesianInferenceMetabolic2018 and
@xingModelingKineticsLargescale2010.

### 2023-10-27 paper/main.tex
Update paper

\title[]{Hierarchical Bayesian regression modelling predicts biochemical
rection energies and facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Below, we review the existing options in this
area, explain how Maud improves on the state of the art, describe the
intended modelling workflow and illustrate its use with an example
application.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, frame hidden, interior hidden]}{\end{tcolorbox}}\fi

@@

In the context of kinetic modelling, Bayesian inference is appealing
because it allows uncertainty to be represented appropriately without
sacrificing mechanistic accuracy. Measurement uncertainty can naturally
be represented in a Bayesian measurement model, whereas the prior model
can represent quantitative uncertainty about kinetic parameters.
Finally, kinetic rate laws can be represented in Bayesian data
generation models with arbitrarily high fidelity. See (REFERENCES) for
more about Bayesian inference and (REFERENCE) for a discussion of
practical Bayesian workflow.

@@

The simulated dataset and underlying kinetic model that we used for our
analysis can be found in {[}\#Supplementary Info{]}.

@@

We used these parameters to simulate steady states in a range of
plausible experimental conditions, again using \citet{saa_general_2015}
as a starting point. These steady states were then used to generate
simulated measurements using the measurement model described below.

\begin{equation}
a = b \label{eq-meth-ode}
\end{equation}

@@

We constructed inputs in Maud's format for each of the analysed
datasets, based on the scenario that the true kinetic model was known
except for parameter values, which needed to be inferred from the
training data and priors. These inputs can be found at XXXX.

The prior distributions and corresponding true parameter values used in
our case study are shown in {[}Supplementary{]}. The first two columns
show the 1\% and 99\% quantiles of each marginal prior distribution.
True parameter value are shown in column three, and the last column
shows the z-score on log scale of the true parameter value according the
marginal prior distribution. As seen in table 2 {[}REFERENCE{]} there
are 7 parameters for which the true value is outside the 1\%-99\% range.
This is desirable, making the case study more realistic, because extreme
deviance from the prior distribution is likely to occur in practice due
to in vivo to in vitro measurement differences.

@@

Overall, our results show that Maud can fit a realistic pathway-sized
dataset. This was achieved without fixing the marginal values of kinetic
parameters: the information required to make good predictions was
contained in the correlation structure of the joint posterior
distribution. This finding is consistent with previous analyses of
biological systems that found they are ``sloppy'', that is, sensitive to
parameter combinations rather than marginal parameter values, with
important combinations, scales and regions of sensitivity being
difficult to ascertain in advance \citet{gutenkunst_2007},
\citet{poirier_revising_1998}.

The question naturally arises whether the crucial high-dimensional
parameter correlations are linear or non-linear. This is relevant to the
question of model performance, as linear correlations are easier to
correct for. A linearly correlated posterior space would also be easier
to summarise. We address this question below in section XXX.

This case study illustrates the type of kinetic model and dataset that
Maud can fit. The model we analysed has XXX reactions, XXX state
variables and XXX parameters. Posterior sampling with adaptive
Hamiltonian Monte Carlo generated XXX ESS/hour. Generalising from this
result, we conclude that it is feasible to use this method to fit models
in the same order of magnitude, but not, for example, genome-scale
kinetic models. To fit larger models, faster steady state solving
methods or alternative inference algorithms will be required. Section
XXX addresses whether Laplace approximation is a suitable candidate.

\hypertarget{comparison-with-laplace-approximation}{%
\subsubsection{Comparison with Laplace
approximation}\label{comparison-with-laplace-approximation}}

@@

The Laplace approximation yielded XXX samples in XXX time. The
diagnostics indicated that our algorithm was able to find the maximum a
posteriori parameter configuration, approximate the Hessian and use
these quantities to generate approximate posterior samples. The results
can be found at XXX.

@@

Maud is a command line application implementing Bayesian inference for a
wide range of realistic kinetic models. Maud is written in Python
\citep{vanrossumPythonReferenceManual2009}, designed for use on Windows,
macOS and Linux, registered on the Python Package Index as
\texttt{maud-metabolic-models}, documented at \textless https://maud-
metabolic-models.readthedocs.io\textgreater{} and actively developed and
maintained at \url{https://github.com/biosustain/Maud/}.

@@

Maud inputs are structured directories, somewhat inspired by the PEtab
format \citet{SchmiesterSch2021}. A Maud input directory must contain a
toml \citep{preston-wernertomandgedampradyunTOMLSpecification0rc2020}
file called \texttt{config.toml} which gives the input a name,
configures how Maud will be run and tells Maud where to find the other
files, allowing these to have custom names. It must also include a file
containing a kinetic model definition, a file specifying information
about parameters and a file with information experiments. The required
structure of these files is documented at \textless https://
maud-metabolic-models.readthedocs.io/en/latest/inputting.html\textgreater.
The input is validated against a Pydantic
\citep{pydanticdevelopersPydantic2022} data model which can be inspected
at \url{https://github.com/biosustain/Maud/tree/main/maud/data_model}.

@@

Each of the terms on the right-hand side of (REFER TO EQUATION) is a
function of \(C\) and \(\theta\). This idea is taken from
\citet{noor_note_2013}. The terms usefully gather physically meaningful
and conceptually distinct factors contributing to reaction fluxes.
\(Enzyme\) captures the effect of enzyme concentration, \(k_{cat}\) that
of enzyme efficiency, \(Reversibility\) quantifies thermodynamic
effects, \(Saturation\) the effect of enzyme availability and
\(Allostery\) the effect of post translational modifications.

@@

Maud's prior model includes unknown parameters corresponding to
quantities in the kinetic model that are assumed to be unknown, other
than steady state metabolite concentrations and fluxes, which are
derived from the values of other parameters by solving the steady state
problem. See Table 1 (REFERENCE) for a description of all these
parameters and their dimensions. Note that some quantities in Maud's
kinetic model are not treated as parameters: for example, temperatures,
compartment volumes and the formation energy of water. Maud treats these
quantities as if they were known precisely: they can be configured by
the user or default values can be used. Although in practice there can
be considerable uncertainty regarding these quantities, we chose to
disregard this uncertainty in the interest of simplicity.

@@

\hypertarget{measurement-model-1}{%
\subsubsection{Measurement model}\label{measurement-model-1}}

@@

\begin{equation}
y_i^{conc} \sim LN(\ln{\hat{y}_i^{conc}}, \sigma_i^{conc})\label{eq-yconc}
\end{equation}

\begin{equation}
y_i^{enz} \sim LN(\ln{\hat{y}_i^{enz}}, \sigma_i^{enz})\label{eq-yenz}
\end{equation}

Both equations are log-normal generalised linear models with a standard
link function (the natural logarithm \(\ln\)) and known standard
deviation \$ \sigma\_\{𝑐𝑜𝑛𝑐\}\$. The use of this measurement model is
motivated by the consideration that concentrations are constrained to be
non-negative, so the measurement model should avoid assigning positive
probability mass to negative metabolite concentration values. In
addition, we expect the precision of most metabolomics and proteomics
experiments to be roughly proportional to the value of the true measured
quantity, which supports a measurement model with constant coefficient
of variation. The measurement standard deviations \(\sigma_{𝑐𝑜𝑛𝑐}\) and
\(\sigma_{𝑒𝑛𝑧}\) are assumed to be known exactly for simplicity;
plausible values can be elicited by considering the likely coefficient
of variation of the measuring apparatus.

@@

Finally, the reader may wonder why Maud uses a linear regression model
for reaction flux measurements even though this creates the potential
for erroneous double counting and requires non-trivial upstream
modelling, as intracellular fluxes typically cannot be measured
directly, but must be inferred from isotope labelling experiments using
metabolic flux analysis (see REFERENCE). Ideally Maud's measurement
model for fluxes would extend from fluxes to the results of potential
labelling experiments, thereby removing the need for upstream analysis
and avoiding any double counting. This option has not yet been pursued,
again for the sake of simplicity.

@@

Although integrals of the joint probability model for kinetic models are
typically analytically intractable, they can be approximated numerically
using Markov Chain Monte Carlo (MCMC) and other methods. Maud uses MCMC
primarily because there exist many methods for verifying that MCMC
samples really do approximate the target probability distribution:
see@vehtariRankNormalizationFoldingLocalization2021 and
\citet{taltsValidatingBayesianInference2018} for discussion of this
point. In addition, there are several examples of successful Bayesian
kinetic modelling projects using MCMC including
\citet{st.johnBayesianInferenceMetabolic2018} and
\citet{xingModelingKineticsLargescale2010}.

>>>>

\title[]{Hierarchical Bayesian regression modelling predicts biochemical
rection energies and facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Below, we review the existing options in this
area, explain how Maud improves on the state of the art, describe the
intended modelling workflow and illustrate its use with an example
application.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, enhanced]}{\end{tcolorbox}}\fi

@@

In the context of kinetic modelling, Bayesian inference is appealing
because it allows uncertainty to be represented appropriately without
sacrificing mechanistic accuracy. Measurement uncertainty can naturally
be represented in a Bayesian measurement model, whereas the prior model
can represent quantitative uncertainty about kinetic parameters.
Finally, kinetic rate laws can be represented in Bayesian data
generation models with arbitrarily high fidelity. See
\citet{gelmanBayesianDataAnalysis2020a} for more about Bayesian
inference and \citet{gelmanBayesianWorkflow2020} for a discussion of
practical Bayesian workflow.

@@

The simulated dataset and underlying kinetic model that we used for our
analysis can be found at
\url{https://github.com/biosustain/Methionine_model/tree/main/data/methionine}
and is described in supplementary information section {[}REFERENCE{]}.

@@

We used these parameters to simulate steady states in a range of
plausible experimental conditions, again using \citet{saa_general_2015}
as a starting point. These steady states were then used to generate
simulated measurements using the measurement model described below in
Section~\ref{sec-methionine-measurement-model}.

\hypertarget{sec-methionine-measurement-model}{%
\subsubsection{Measurement
model}\label{sec-methionine-measurement-model}}

@@

We constructed inputs in Maud's format for each of the analysed
datasets, based on the scenario that the true kinetic model was known
except for parameter values, which needed to be inferred from the
training data and priors. These inputs can be found at
\url{https://github.com/biosustain/Methionine_model/tree/main/data}.

The prior distributions and corresponding true parameter values used in
our case study are shown in supplementary materials section
{[}REFERENCE{]}. The first two columns show the 1\% and 99\% quantiles
of each marginal prior distribution. True parameter value are shown in
column three, and the last column shows the z-score on log scale of the
true parameter value according the marginal prior distribution. As can
be seen from the table, there are 7 parameters for which the true value
is outside the 1\%-99\% range. This is desirable, making the case study
more realistic, because extreme deviance from the prior distribution is
likely to occur in practice due to in vivo to in vitro measurement
differences.

@@

Overall, our results show that Maud can fit a realistic pathway-sized
dataset. This was achieved without fixing the marginal values of kinetic
parameters: the information required to make good predictions was
contained in the correlation structure of the joint posterior
distribution. This finding is consistent with previous analyses of
biological systems that found they are ``sloppy'', that is, sensitive to
parameter combinations rather than marginal parameter values, with
important combinations, scales and regions of sensitivity being
difficult to ascertain in advance
\citep{gutenkunst_2007, poirier_revising_1998}.

The question naturally arises whether the crucial high-dimensional
parameter correlations are linear or non-linear. This is relevant to the
question of model performance, as linear correlations are easier to
correct for. A linearly correlated posterior space would also be easier
to summarise. We address this question below in
Section~\ref{sec-laplace}.

This case study illustrates the type of kinetic model and dataset that
Maud can fit. The model we analysed has XXX reactions, XXX state
variables and XXX parameters. Posterior sampling with adaptive
Hamiltonian Monte Carlo generated XXX ESS/hour. Generalising from this
result, we conclude that it is feasible to use this method to fit models
in the same order of magnitude, but not, for example, genome-scale
kinetic models. To fit larger models, faster steady state solving
methods or alternative inference algorithms will be required.
Section~\ref{sec-laplace} addresses whether Laplace approximation is a
suitable candidate.

\hypertarget{sec-laplace}{%
\subsubsection{Comparison with Laplace
approximation}\label{sec-laplace}}

@@

The Laplace approximation yielded XXX samples in XXX time. The
diagnostics indicated that our algorithm was able to find the maximum a
posteriori parameter configuration, approximate the Hessian and use
these quantities to generate approximate posterior samples. The results
can be found at
\url{https://github.com/biosustain/Methionine_model/tree/main/results}.

@@

Maud is a command line application implementing Bayesian inference for a
wide range of realistic kinetic models. Maud is written in Python
\citep{vanrossumPythonReferenceManual2009}, designed for use on Windows,
macOS and Linux, registered on the Python Package Index as
\texttt{maud-metabolic-models}, documented at
\url{https://maud-metabolic-models.readthedocs.io} and actively
developed and maintained at \url{https://github.com/biosustain/Maud/}.

@@

Maud inputs are structured directories, somewhat inspired by the PEtab
format \citep{SchmiesterSch2021}. A Maud input directory must contain a
toml \citep{preston-wernertomandgedampradyunTOMLSpecification0rc2020}
file called \texttt{config.toml} which gives the input a name,
configures how Maud will be run and tells Maud where to find the other
files, allowing these to have custom names. It must also include a file
containing a kinetic model definition, a file specifying information
about parameters and a file with information experiments. The required
structure of these files is documented at \textless https://
maud-metabolic-models.readthedocs.io/en/latest/inputting.html\textgreater.
The input is validated against a Pydantic
\citep{pydanticdevelopersPydantic2022} data model which can be inspected
at \url{https://github.com/biosustain/Maud/tree/main/maud/data_model}.

@@

Each of the terms on the right-hand side of \eqref{eq-decomposition} is
a function of \(C\) and \(\theta\). This idea is taken from
\citet{noor_note_2013}. The terms usefully gather physically meaningful
and conceptually distinct factors contributing to reaction fluxes.
\(Enzyme\) captures the effect of enzyme concentration, \(k_{cat}\) that
of enzyme efficiency, \(Reversibility\) quantifies thermodynamic
effects, \(Saturation\) the effect of enzyme availability and
\(Allostery\) the effect of post translational modifications.

@@

Maud's prior model includes unknown parameters corresponding to
quantities in the kinetic model that are assumed to be unknown, other
than steady state metabolite concentrations and fluxes, which are
derived from the values of other parameters by solving the steady state
problem. See table (REFERENCE) in this paper's supplementary materials
for a description of all these parameters and their dimensions. Note
that some quantities in Maud's kinetic model are not treated as
parameters: for example, temperatures, compartment volumes and the
formation energy of water. Maud treats these quantities as if they were
known precisely: they can be configured by the user or default values
can be used. Although in practice there can be considerable uncertainty
regarding these quantities, we chose to disregard this uncertainty in
the interest of simplicity.

@@

@@

\begin{align}
y_i^{conc} &\sim LN(\ln{\hat{y}_i^{conc}}, \sigma_i^{conc})\label{eq-yconc} \\
y_i^{enz} &\sim LN(\ln{\hat{y}_i^{enz}}, \sigma_i^{enz})\label{eq-yenz}
\end{align}

Both equations are log-normal generalised linear models with a standard
link function (the natural logarithm \(\ln\)) and known standard
deviation \(\sigma_{𝑐𝑜𝑛𝑐}\). The use of this measurement model is
motivated by the consideration that concentrations are constrained to be
non-negative, so the measurement model should avoid assigning positive
probability mass to negative metabolite concentration values. In
addition, we expect the precision of most metabolomics and proteomics
experiments to be roughly proportional to the value of the true measured
quantity, which supports a measurement model with constant coefficient
of variation. The measurement standard deviations \(\sigma_{𝑐𝑜𝑛𝑐}\) and
\(\sigma_{𝑒𝑛𝑧}\) are assumed to be known exactly for simplicity;
plausible values can be elicited by considering the likely coefficient
of variation of the measuring apparatus.

@@

Finally, the reader may wonder why Maud uses a linear regression model
for reaction flux measurements even though this creates the potential
for erroneous double counting and requires non-trivial upstream
modelling, as intracellular fluxes typically cannot be measured
directly. Instead, fluxes are typically inferred from isotope labelling
experiments using metabolic flux analysis: see
\citet{daiUnderstandingMetabolismFlux2017} for more about this method.
Ideally Maud's measurement model for fluxes would extend from fluxes to
the results of potential labelling experiments, thereby removing the
need for upstream analysis and avoiding any double counting. This option
has not yet been pursued, again for the sake of simplicity.

@@

Although integrals of the joint probability model for kinetic models are
typically analytically intractable, they can be approximated numerically
using Markov Chain Monte Carlo (MCMC) and other methods. Maud uses MCMC
primarily because there exist many methods for verifying that MCMC
samples really do approximate the target probability distribution: see
\citet{vehtariRankNormalizationFoldingLocalization2021} and
\citet{taltsValidatingBayesianInference2018} for discussion of this
point. In addition, there are several examples of successful Bayesian
kinetic modelling projects using MCMC including
\citet{st.johnBayesianInferenceMetabolic2018} and
\citet{xingModelingKineticsLargescale2010}.
